The Impact of Malaria Episodes and Treatment Regimens on Adverse Pregnancy Outcomes in Ugandan Women

### 2025-03-12 code/eda-code/eda_MADAprojectcode.qmd
Project Update

```{r}
# Load required packages
library(dplyr)
library(gtsummary)

# Select relevant variables for malaria episodes and birth outcomes
malaria_outcomes_table <- PROMO_Data %>%
  select(study_arm, 
         total_malaria_episodes_during_pregnancy, 
         placental_malaria, 
         preterm_birth_ontoneo, 
         stillbirth, 
         infant_low_birth_weight_2500_g) %>%
  tbl_summary(
    by = study_arm,  # Compare DP vs. SP
    statistic = list(
      total_malaria_episodes_during_pregnancy ~ "{mean} ± {sd}",
      all_categorical() ~ "{n} ({p}%)"
    ),
    label = list(
      total_malaria_episodes_during_pregnancy ~ "Total Malaria Episodes (Mean ± SD)",
      placental_malaria ~ "Placental Malaria (Histology)",
      preterm_birth_ontoneo ~ "Preterm Birth",
      stillbirth ~ "Stillbirth",
      infant_low_birth_weight_2500_g ~ "Low Birth Weight (<2500g)"
    )
  ) %>%
  add_p() %>%
  bold_labels()

# Print the formatted table
malaria_outcomes_table

@@

*Categorical Characteristics by Study Arm*

```{r}
# Generate a summary table for categorical characteristics
categorical_table <- PROMO_Data %>%
  select(study_arm, stillbirth, neonatal_death, infant_low_birth_weight_2500_g) %>%
  tbl_summary(by = study_arm, 
              statistic = list(all_categorical() ~ "{n} ({p}%)"),
              label = list(stillbirth ~ "Stillbirth",
                           neonatal_death ~ "Neonatal Death",
                           infant_low_birth_weight_2500_g ~ "Low Birth Weight")) %>%
  add_p() %>%
  bold_labels()

categorical_table

@@

```{r}
ggplot(PROMO_Data, aes(x = total_malaria_episodes_during_pregnancy)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Malaria Episodes", x = "Malaria Episodes", y = "Count") +
  theme_minimal()

>>>>

```{r}
colnames(PROMO_Data)

@@

```{r}
# Load required libraries
# Load required libraries
library(dplyr)
library(gtsummary)

# Define malaria outcome variables
malaria_outcome_vars <- c(
  "total_malaria_episodes_during_pregnancy",
  "placental_malaria",
  "preterm_birth_ontoneo",
  "stillbirth",
  "infant_low_birth_weight_2500_g"
)

# Ensure that `study_arm` is in the dataset
if (!"study_arm" %in% colnames(PROMO_Data)) {
  stop("⚠️ Error: `study_arm` column not found in PROMO_Data. Please check column names.")
}

# Remove `stillbirth` and any other variable that has only one level
valid_outcomes <- malaria_outcome_vars[sapply(PROMO_Data[malaria_outcome_vars], function(x) length(unique(na.omit(x))) > 1)]

# Ensure `study_arm` is included in the selected variables
valid_outcomes <- c(valid_outcomes, "study_arm")

# Define label mappings dynamically based on valid variables
all_labels <- list(
  total_malaria_episodes_during_pregnancy ~ "Total Malaria Episodes (Mean ± SD)",
  placental_malaria ~ "Placental Malaria (Histology)",
  preterm_birth_ontoneo ~ "Preterm Birth",
  stillbirth ~ "Stillbirth",
  infant_low_birth_weight_2500_g ~ "Low Birth Weight (<2500g)"
)

dynamic_labels <- all_labels[names(all_labels) %in% valid_outcomes]

# Define a custom function to format p-values
format_pvalue <- function(pvals) {
  sapply(pvals, function(x) {
    if (is.na(x)) {
      return("NA")  # Handle missing values properly
    } else if (is.numeric(x)) {
      if (x < 0.001) {
        return("< 0.001")  # Report very small values as "< 0.001"
      } else if (x < 0.01) {
        return(format(round(x, 3), nsmall = 3))  # Report between 0.001 and 0.01 to three decimal places
      } else {
        return(format(round(x, 2), nsmall = 2))  # Report >= 0.01 to two decimal places
      }
    } else {
      return("NA")  # Return "NA" for non-numeric values
    }
  })
}

# Generate table with formatted p-values
if (nrow(PROMO_Data) > 0 && length(valid_outcomes) > 1) {

  malaria_outcomes_table <- PROMO_Data %>%
    select(all_of(valid_outcomes)) %>%
    tbl_summary(
      by = "study_arm",
      statistic = list(
        total_malaria_episodes_during_pregnancy ~ "{mean} ± {sd}",
        all_categorical() ~ "{n} ({p}%)"
      ),
      label = dynamic_labels
    ) %>%
    add_p(
      test = list(
        all_categorical() ~ "fisher.test",
        all_continuous() ~ "t.test"
      ),
      pvalue_fun = format_pvalue  # Apply the fixed p-value formatting function
    )

  # Print the updated table
  print(malaria_outcomes_table)

} else {
  message("⚠️ No data available for analysis or insufficient variability in variables.")
}
```

@@

```{r}
ggplot(PROMO_Data, aes(x = total_malaria_episodes_during_pregnancy)) +
  geom_histogram(binwidth = 1, fill = "light blue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Malaria Episodes During Pregnancy", 
       x = "Number of Malaria Episodes", 
       y = "Frequency") +
  theme_minimal()

*Histogram of Gestational Age at Delivery*
```{r}
ggplot(PROMO_Data, aes(x = gestational_age_at_delivery_weeks)) +
  geom_histogram(binwidth = 1, fill = "light blue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Gestational Age at Delivery", 
       x = "Gestational Age (weeks)", 
       y = "Number of Births") +
  theme_minimal()

*Histogram of Infant Birth Weight*
```{r}
ggplot(PROMO_Data, aes(x = infant_birth_weight_kg)) +
  geom_histogram(binwidth = 0.2, fill = "green", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Infant Birth Weights", 
       x = "Birth Weight (kg)", 
       y = "Number of Infants") +
  theme_minimal()

*Histogram of Maternal Age*
```{r}
ggplot(PROMO_Data, aes(x = age_at_enrollment_years)) +
  geom_histogram(binwidth = 2, fill = "light blue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Maternal Age at Enrollment", 
       x = "Maternal Age (years)", 
       y = "Number of Participants") +
  theme_minimal()

*Histogram of Malaria Episodes Stratified by Treatment Regimen (SP vs DP)*
```{r}
ggplot(PROMO_Data, aes(x = total_malaria_episodes_during_pregnancy, fill = study_arm)) +
  geom_histogram(binwidth = 1, color = "black", alpha = 0.7, position = "identity") +
  facet_wrap(~ study_arm) +
  labs(title = "Malaria Episodes Stratified by Treatment Regimen", 
       x = "Number of Malaria Episodes", 
       y = "Frequency") +
  theme_minimal()

### 2025-03-12 products/manuscript/Manuscript.qmd
Project Update

---
knitr:
  opts_chunk:
    echo: false
    message: false
    warning: false

title: "The Impact of Malaria Episodes and Treatment Regimens on Adverse Pregnancy Outcomes in Ugandan Women"
author: "Asmith Joseph"
format:
  docx:
    toc: false
    number-sections: true
    reference-location: margin
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Load Required Libraries
# Essential libraries for data handling, visualization, and analysis

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Define the correct file path using the `here` package
data_path <- here("data", "raw-data", "PROMO_Data.csv")

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Data Import & Initial Inspection

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#  Data Cleaning & Transformation

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#  Load required libraries
library(dplyr)
library(gtsummary)

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Load required libraries
library(dplyr)
library(gtsummary)

# Define key variables (check if they exist in dataset)
malaria_outcome_vars <- c("study_arm", 
                           "total_malaria_episodes_during_pregnancy", 
                           "placental_malaria", 
                           "preterm_birth_ontoneo", 
                           "stillbirth", 
                           "infant_low_birth_weight_2500_g")

#  Keep only variables that exist in dataset
malaria_outcome_vars <- intersect(malaria_outcome_vars, colnames(PROMO_Data))

#  Generate the table only if dataset has rows
if (nrow(PROMO_Data) > 0) {
  malaria_outcomes_table <- PROMO_Data %>%
    select(all_of(malaria_outcome_vars)) %>%
    tbl_summary(
      by = study_arm,  # Compare DP vs. SP groups
      statistic = list(
        total_malaria_episodes_during_pregnancy ~ "{mean} ± {sd}",  # Show mean ± SD
        all_categorical() ~ "{n} ({p}%)"  # Show counts and percentages for categorical variables
      ),
      label = list(
        total_malaria_episodes_during_pregnancy ~ "Total Malaria Episodes (Mean ± SD)",
        placental_malaria ~ "Placental Malaria (Histology)",
        preterm_birth_ontoneo ~ "Preterm Birth",
        stillbirth ~ "Stillbirth",
        infant_low_birth_weight_2500_g ~ "Low Birth Weight (<2500g)"
      )
    ) %>%
    add_p() %>%  # Add p-values for comparisons
    modify_header(label = "**Variable**") %>%  # Ensure readable column headers
    bold_labels()  # Bold table labels for clarity

  #  Display only the formatted table
  malaria_outcomes_table
} else {
  cat("️ No data available for analysis.")
}

>>>>

---
title: "The Impact of Malaria Episodes and Treatment Regimens on Adverse Pregnancy Outcomes in Ugandan Women"
author: "Asmith Joseph"  
format:
  docx:
    toc: false
    number-sections: true
    reference-location: margin
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
knitr:
  opts_chunk:
    echo: false
    message: false
    warning: false
---

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold"}
# Load Required Libraries
# Essential libraries for data handling, visualization, and analysis

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold"}
# Define the correct file path using the `here` package
data_path <- here("data", "raw-data", "PROMO_Data.csv")

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold"}
# Data Import & Initial Inspection

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold"}
#  Data Cleaning & Transformation

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold"}
#  Load required libraries
library(dplyr)
library(gtsummary)

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold"}
# Load required libraries
library(dplyr)
library(gtsummary)

# Define malaria outcome variables
malaria_outcome_vars <- c(
  "total_malaria_episodes_during_pregnancy",
  "placental_malaria",
  "preterm_birth_ontoneo",
  "stillbirth",
  "infant_low_birth_weight_2500_g"
)

# Ensure that `study_arm` is in the dataset
if (!"study_arm" %in% colnames(PROMO_Data)) {
  stop("⚠️ Error: `study_arm` column not found in PROMO_Data. Please check column names.")
}

# Remove `stillbirth` and any other variable that has only one level
valid_outcomes <- malaria_outcome_vars[sapply(PROMO_Data[malaria_outcome_vars], function(x) length(unique(na.omit(x))) > 1)]

# Ensure `study_arm` is included in the selected variables
valid_outcomes <- c(valid_outcomes, "study_arm")

# Define label mappings dynamically based on valid variables
all_labels <- list(
  total_malaria_episodes_during_pregnancy ~ "Total Malaria Episodes (Mean ± SD)",
  placental_malaria ~ "Placental Malaria (Histology)",
  preterm_birth_ontoneo ~ "Preterm Birth",
  stillbirth ~ "Stillbirth",
  infant_low_birth_weight_2500_g ~ "Low Birth Weight (<2500g)"
)

dynamic_labels <- all_labels[names(all_labels) %in% valid_outcomes]

# Define a custom function to format p-values
format_pvalue <- function(pvals) {
  sapply(pvals, function(x) {
    if (is.na(x)) {
      return("NA")  # Handle missing values properly
    } else if (is.numeric(x)) {
      if (x < 0.001) {
        return("< 0.001")  # Report very small values as "< 0.001"
      } else if (x < 0.01) {
        return(format(round(x, 3), nsmall = 3))  # Report between 0.001 and 0.01 to three decimal places
      } else {
        return(format(round(x, 2), nsmall = 2))  # Report >= 0.01 to two decimal places
      }
    } else {
      return("NA")  # Return "NA" for non-numeric values
    }
  })
}

# Generate table with formatted p-values
if (nrow(PROMO_Data) > 0 && length(valid_outcomes) > 1) {

  malaria_outcomes_table <- PROMO_Data %>%
    select(all_of(valid_outcomes)) %>%
    tbl_summary(
      by = "study_arm",
      statistic = list(
        total_malaria_episodes_during_pregnancy ~ "{mean} ± {sd}",
        all_categorical() ~ "{n} ({p}%)"
      ),
      label = dynamic_labels
    ) %>%
    add_p(
      test = list(
        all_categorical() ~ "fisher.test",
        all_continuous() ~ "t.test"
      ),
      pvalue_fun = format_pvalue  # Apply the fixed p-value formatting function
    )

  # Print the updated table
  print(malaria_outcomes_table)

} else {
  message("⚠️ No data available for analysis or insufficient variability in variables.")
}
